A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market

### 2013-08-29 reports/Draft.Rnw
resolved conflict
working on lit review section

\section{Introduction and Past Research}\label{sec:lit}
Prolonged exposure to traffic noise affects people in a number of ways, ranging from simple annoyance (Miedema and Oudshoorn 2001; Ouis 2001), to sleep disturbance (Netherlands 2004), to increasing risk for stroke (Sorensen, Hvidberg et al. 2011), hypertension(Jarup, Babishch et al. 2008; Bodin, Albin et al. 2009), and myocardial infarction (Babisch, Beule et al. 2005). The noise level at which such effects are observed does not have to be high.  It has been shown that people exposed to traffic noise with a 24-hour average of 55 dBA are found to be at a higher risk for hypertension (Barregard, Bonde et al. 2009; Bodin, Albin et al. 2009), and those exposed to 60 dBA or greater are found to be at a higher risk for stroke (Sorensen, Hvidberg et al. 2011).  This impact is expected to grow worse with the increasing number of vehicles in the urban road network and the diminishing number of night-time quiet hours unless mitigating measures are put in place (XXX).  Given that much of this increased exposure occurs in areas where people live and work any mitigating measures must be based on a cost-benefit analysis.

An important first step towards such an analysis is to examine real estate price fluctuations as a function of noise levels using hedonic regression.  Such work has provided important insight on the nature of this relationship.  Thus, \citet{Bateman2001} found that the negative eﬀect of traﬃc noise ranged from 0.08 to 2.22 percent per dB with a mean around 0.55 percent per dB. In a review of 65 noise evaluation studies \citet{Navrud2002} found an average decrease of 0.64 percent per dB. \textbf{We need to briefly review other relevant papers here.}  However, with the exception of a few studies that have focused on air traffic noise exposure \textbf{(see, for instance, \citet{coh})}, the relationship between road traffic noise levels and real estate prices has rarely been explored in the United States, even though the former is perhaps the greatest source of noise in residential neighborhoods (XXX).  One major reason for the thinness of the scholarly literature on this topic in the US lies in the difficulty of modeling noise propagation over the landscape.  
>>>>>>> 4543f31da98e1080a1ead85b6607230c99f3c848

>>>>

\section{Introduction and Past Research}\label{sec:lit}
Prolonged exposure to traffic noise affects people in a number of ways, ranging from simple annoyance \citep{Weinhold2013, Miedema2001} (Miedema and Oudshoorn 2001; Ouis 2001), to sleep disturbance (Netherlands 2004), to increasing risk for stroke (Sorensen, Hvidberg et al. 2011), hypertension(Jarup, Babishch et al. 2008; Bodin, Albin et al. 2009), and myocardial infarction (Babisch, Beule et al. 2005). The noise level at which such effects are observed does not have to be high.  It has been shown that people exposed to traffic noise with a 24-hour average of 55 dBA are found to be at a higher risk for hypertension (Barregard, Bonde et al. 2009; Bodin, Albin et al. 2009), and those exposed to 60 dBA or greater are found to be at a higher risk for stroke (Sorensen, Hvidberg et al. 2011).  This impact is expected to grow worse with the increasing number of vehicles in the urban road network and the diminishing number of night-time quiet hours unless mitigating measures are put in place (XXX).  Given that much of this increased exposure occurs in areas where people live and work any mitigating measures must be based on a cost-benefit analysis.

An important first step towards such an analysis is to examine real estate price fluctuations as a function of noise levels using hedonic regression. \citet{Nelson1982} reviewed early work in this area, but many early studies were limited by small sample sizes and narrow spatial scopes. A review by \citet{Bateman2001} found that the negative eﬀect of traﬃc noise ranged from 0.08 to 2.22 percent per dB with a mean around 0.55 percent per dB, while \citet{Navrud2002} found an average decrease of 0.64 percent per dB. More recent work has tended to focus on the European experience with noise, especially the United Kingdom \citep{Day2007, Blanco2011}, Sweden \citep{Wilhelmsson2000, Andersson2010}, Netherlands \citep{Theebe2004a}, Spain \citep{MarmolejoDuarteCarlos;GonzalezTamez2009}, and Switzerland \citep{Baranzini2010}.

Work examining the negative effects of noise in the United States have tended to concentrate on noise from airports \citep{Espey2000, McMillen2004, Cohen2008a} or indirect effects of traffic noise, such as proximity to highways \citep{Matthews2007, Chernobai2009, Li2012}, or traffic counts \citep{Highes1992, Larsen2012}. the relationship between road traffic noise levels and real estate prices has rarely been explored in the United States, even though the former is perhaps the greatest source of noise in residential neighborhoods (XXX).  One major reason for the thinness of the scholarly literature on this topic in the US lies in the difficulty of modeling noise propagation over the landscape.

### 2013-08-29 reports/Draft.tex
working on lit review section

\begin{abstract}
This paper implements the hedonic price method to estimate the impact of traffic noise on properties within the St.\ Paul, Minnesota metropolitan real-estate market over the time period from 2005 - 2010. We use Locally Weighted Regression (LWR) estimation techniques in order to allow the hedonic function to vary over time and space. Significant evidence suggests spatial and temporal non-stationarity in the hedonic price function within our study area. In contrast to previous research, however, we find little evidence that the impact of traffic noise exhibits a significant non-linear relationship with the decibel level of the traffic noise, nor does the impact seem to vary over space. We find significant differences in the impact of traffic noise from the beginning to the end of our sample. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September, 2008, and an average of 0.37 percent after September, 2008. While significantly different, both of these estimates are still within the typical range of previous estimates in the literature.
\end{abstract}

\section{Introduction and Past Research}\label{sec:lit}
Prolonged exposure to traffic noise affects people in a number of ways, ranging from simple annoyance (Miedema and Oudshoorn 2001; Ouis 2001), to sleep disturbance (Netherlands 2004), to increasing risk for stroke (Sorensen, Hvidberg et al. 2011), hypertension(Jarup, Babishch et al. 2008; Bodin, Albin et al. 2009), and myocardial infarction (Babisch, Beule et al. 2005). The noise level at which such effects are observed does not have to be high.  It has been shown that people exposed to traffic noise with a 24-hour average of 55 dBA are found to be at a higher risk for hypertension (Barregard, Bonde et al. 2009; Bodin, Albin et al. 2009), and those exposed to 60 dBA or greater are found to be at a higher risk for stroke (Sorensen, Hvidberg et al. 2011).  This impact is expected to grow worse with the increasing number of vehicles in the urban road network and the diminishing number of night-time quiet hours unless mitigating measures are put in place (XXX).  Given that much of this increased exposure occurs in areas where people live and work any mitigating measures must be based on a cost-benefit analysis.

An important first step towards such an analysis is to examine real estate price fluctuations as a function of noise levels using hedonic regression.  Such work has provided important insight on the nature of this relationship.  Thus, Bateman et al. (2001) found that the negative eﬀect of traﬃc noise ranged from 0.08\% to 2.22\% per dB with a mean around 0.55\% per dB. In a review of 65 noise evaluation studies Navrud (2002) found an average decrease of 0.64\% per dB. \textbf{We need to briefly review other relevant papers here.}  However, with the exception of a few studies that have focused on air traffic noise exposure (we should check on this), the relationship between road traffic noise levels and real estate prices has rarely been explored in the United States, even though the former is perhaps the greatest source of noise in residential neighborhoods (XXX).  One major reason for the thinness of the scholarly literature on this topic in the US lies in the difficulty of modeling noise propagation over the landscape.

To properly analyze the spatial association between real estate prices and exposure to traffic noise at a landscape level, it is necessary to create a noise surface map at sufficiently detailed spatial resolution to account for the complex and heterogeneous interaction between the noise source and the resistance of the landscape to noise propagation.  Implementing such a model can be very difficult.  For one thing, the data needed for the model is very extensive and may not even be readily available (e.g., building footprint and height data).  Furthermore, it is computationally very intensive, requiring more computing resources than a desktop. Fortunately, recent developments in GIS and distributive computing have reduced these difficulties, making it much easier to create a noise surface map at landscape level.

In this paper, we seek to contribute to the hedonic literature by examining the relationship between real estate prices and variation in road traffic noise exposure.  We approach this issue by means of an empirical case study of the Twin Cities Metro Region (TCMR), Minnesota.  In particular, we model traffic noise propagation at landscape level using a spatially explicit approach instead of relying on (what?)…

The value of noise pollution is typically estimated through simulated or related markets because no explicit private market for noise abatement exists. \citet{Nelson1982} is perhaps the best known earliest review of 14 housing market studies of the economic impact of traffic noise on property values in Canada and the United States. More recently, a number of reviews have been produced for European government agencies. \citet{Bateman2001} reviewed the literature for the Scottish Government while \citet{Navrud2002} provided a ``state of the art'' review for the European Commission Environment Directorate General.

The Hedonic Price Method is commonly used to value noise by examining real estate price fluctuations as a function of noise levels. Based on the canonical work by \citet{Rosen1974}, researchers typically estimate a hedonic price function specifying the logged sale price of a property as a linear function of the traffic noise exhibited at the house. Such a specification leads to reporting the impact of traffic noise as a percentage decrease in house price per one decibel (dB) increase in traffic noise. According to \citet{Bateman2001}, the negative effect of traffic noise has ranged from 0.08\% to 2.22\% per dB with a mean around 0.55\% per dB. In a review of 65 noise evaluation studies \citet{Navrud2002} found an average decrease of 0.64\% per dB.

\citet{Nelson2008} produced an updated review of the traffic noise literature for the edited volume \underline{Hedonic Methods in Housing Markets Economics}. The chapter describes several areas of continued interest in the empirical literature, most notably: the measure of traffic noise, spatial modeling of the hedonic price function, and housing market dynamics.

\subsection{Measure of Traffic Noise}
I'm going to need some help writing about noise measures. Perhaps Tsegaye can lend a hand?

Noise data is typically not collected for each every house in a hedonic study. For example, \citet{Huang;Palmquist2001} collect noise data using 128 monitoring stations throughout a suburb of Seattle, WA. Such a strategy forces the researchers to interpolate the data between the monitoring stations in order to associate each house with a noise level. While the researchers do not describe how the interpolation was done, they implicitly admit the potential uncertainty of the method by categorizing traffic noise exposure into 2.5 dB wide categories. Other researchers, such as \citet{Theebe2004a} use 5 dB wide bins, and \citet{Nelson2008} even claims that some studies use bins as wide as 10 dB. Such coarse measurements can lead to serious errors in variables and imprecise estimates of the impact of noise.

Other researchers obtain noise levels from external sources without much description of the variable source. For instance, \cite{MarmolejoDuarteCarlos;GonzalezTamez2009} simply list a ``City Council Sonic Map'',  \citet{Blanco2011} cite the Department for Environment Food and Rural Affairs Noise mapping England project. \citet{Andersson2010} state, ``Information about noise levels is from a study on the health effects of traffic noise conducted in Lerum [Sweden] in 2004 (Ohrstrom et al. 2005). Separate noise calculations were made for railway and road noise for all the houses in Lerum.''

Traffic noise at a location can vary significantly depending on changing traffic conditions. Such variation has led to multiple measures of noise. Some measures eliminate the fluctuations in sound and instead represent the equivalent continuous constant sound pressure at a site. Other indices use noise values that are only exceeded a given percentage of time. For instance, \citet{Huang;Palmquist2001} use dB $L_{10}$ as their noise measure, where $L_{10}$ denotes the value that is exceeded only 10 percent of the time. Still others adjust nighttime noise exposure levels to account for potential increased perceived nuisance. \citet{Baranzini2010} even surveyed inhabitants about their perceptions of the level of noise and their annoyance. The work finds similar results for significant levels of measured noise as compared subjective measures of noise.

\emph{What do we have? There is still a lot of work to be done in this area, better understanding the measures others use as well as what we have and how they compare. I'm surprised at how little information people actually provide about their noise data.}

Spatial resolution of the data. \citet{Theebe2004a} had noise levels for square areas of 100 x 100m.

% Beyond the fact that no direct natural economic markets exist in which consumers can reveal their willingness to pay to avoid traffic noise, researchers also must choose how to measure the nuisance itself. Some researchers (see, for instance, \citet{HughesJr.1992}, \citet{Li2012} \citet{Larsen2012}) attempt to measure the negative externality of traffic by correlating traffic volume with house prices. However, traffic may generate other externalities beyond noise (such as local air pollution and reduced safety for pedestrians and drivers) and so these strategies overestimate the size of the noise externality. 
% 
% The value of traffic noise is important in and of itself. Governments routinely erect noise barriers to shield residential areas from traffic noise. (cite some numbers here!). To aid such work, it is important to understand the benefits of these policies. For instance, until 2011, the Minnesota Department of Transportation used a cost effectiveness threshold for noise mitigation projects of \$3,250 per dB reduced per house. Given that some of the benefits of noise reductions are capitalized in house prices, we are curious to see how our estimated magnitudes compare to the state cost effectiveness cutoff.

\subsection{Spatial Analysis}
Advances in Geographic Information System (GIS) technology and digitized datasets have made research easier to conduct over larger geographies, with larger samples, and while explicitly considering the spatial nature of the housing market. Estimation of the hedonic function using Ordinarly Least Squares (OLS) is typically inappropriate because the housing data commonly exhibit spatial autocorrelation. In particular, spatially autocorrelated error terms will lead to inconsistent estimates of OLS regression standard errors and inappropriate measures of statistical inference.

It is now common to use spatial econometric techniques in the hedonic literature, for instance estimating a spatial lag (in which house prices are modeled as a function of neighboring house prices), or a spatial error model (in which the degree of spatial autocorrelation is also estimated). In the noise literature, for example, \citet{Theebe2004a} econometrically estimates a spatial error model, while \citet{Andersson2010} uses a spatial lag model.

One common reason for significant spatially autocorrelated error terms is hedonic function misspecification. A portion of the hedonic regression errors may be explained by omitted spatial variables. Thus, nearby houses will tend to have similar spatial variable values and then similar error terms. Second, model misspecification can occur due to spatial non-stationarity of the hedonic function. That is, housing submarkets exist and the impact of changing housing characteristics will vary across submarkets. Failing to properly account for such heterogeneity will lead to spatially autocorrelated error estimates.

Some traffic noise research simply explicitly assumes that their data comprise a single market. For example, \citet{Kim2007} assumes that by restricting analysis to residential properties surrounding the Inner Circular highway in the central area of Seoul from 2002 to 2004, that the price of traffic noise will be constant. Other research, such as \citet{Theebe2004a} used political boundaries to segment the market into five separate subarkets in the western part of the Netherlands.

Another strategy is to let the data reveal different relationships within smaller areas of the study. For instance, \citet{Day2007} and \citet{MarmolejoDuarteCarlos;GonzalezTamez2009} use semi-parametric techniques to allow the hedonic relationship to smoothly vary over space. We use techniques similar to the Geographically Weighted Regression method described in \citet{MarmolejoDuarteCarlos;GonzalezTamez2009} to allow the hedonic function to vary over space without having a priori conceptions of where the function will be different.

\subsection{Temporal Variation}
Little work has been done estimating the changes in hedonic environmental (dis)amenities like noise over time. An event-study literature exists using temporal variation in noise, especially noise from airports (see, for example, \citet{Cohen2009}, ). However, these articles usually use the change in noise over time to identify the hedonic price of noise, rather than ask whether the hedonic price has changed.

\citet{Wilhelmsson2000} is one example of research asking whether the implicit price of noise has changed over time. The study splits the 290 Swedish house sales into two separate time periods: 1986-89 and 1990-95. A Chow test revealed that the impact of traffic noise was different in the earlier vs. later time period, with noise being considered more of a nuisance in the suburb of Stockholm during the 1990s.

\citet{Cho2011b} is a recent example of work asking whether the hedonic regression coefficients change over time in the Nashville, TN area. In particular, the study uses a repeat sales methodology to estimate changes in the implicit prices of hedonic characteristics as a result of the Great Recession of 2008-09 in the United States. The work finds that environmental amenities matter less in sales during the year 2008 when compared to pre-recession sales.

We estimate regression coefficients over time using a moving window of 12 months in the hedonic analysis over the years 2005 - 2010. Such an analysis allows us to see if and when the implicit price of traffic noise changes within our study timeframe. Our results suggest that the percentage change in house price from an extra decibel of traffic noise actually became more negative after the Great Recession than before.

>>>>

\begin{abstract}
This paper implements the hedonic price method to estimate the impact of traffic noise on properties within the St.\ Paul, Minnesota metropolitan real-estate market over the time period from 2005 - 2010. We use Locally Weighted Regression (LWR) estimation techniques in order to allow the hedonic function to vary over time and space. Significant evidence suggests spatial and temporal non-stationarity in the hedonic price function within our study area. In contrast to previous research, however, we find little evidence that the price semi-elasticity of traffic noise varies with the decibel level of the traffic noise, nor does the impact seem to exhibit spatial non-stationarity. We find significant differences in the impact of traffic noise from the beginning to the end of our sample. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September, 2008, and an average of 0.37 percent after September, 2008. While significantly different, both of these estimates are still within the typical range of previous estimates in the literature.
\end{abstract}

\section{Introduction and Past Research}\label{sec:lit}
Prolonged exposure to traffic noise affects people in a number of ways, ranging from simple annoyance \citep{Weinhold2013, Miedema2001} (Miedema and Oudshoorn 2001; Ouis 2001), to sleep disturbance (Netherlands 2004), to increasing risk for stroke (Sorensen, Hvidberg et al. 2011), hypertension(Jarup, Babishch et al. 2008; Bodin, Albin et al. 2009), and myocardial infarction (Babisch, Beule et al. 2005). The noise level at which such effects are observed does not have to be high.  It has been shown that people exposed to traffic noise with a 24-hour average of 55 dBA are found to be at a higher risk for hypertension (Barregard, Bonde et al. 2009; Bodin, Albin et al. 2009), and those exposed to 60 dBA or greater are found to be at a higher risk for stroke (Sorensen, Hvidberg et al. 2011).  This impact is expected to grow worse with the increasing number of vehicles in the urban road network and the diminishing number of night-time quiet hours unless mitigating measures are put in place (XXX).  Given that much of this increased exposure occurs in areas where people live and work any mitigating measures must be based on a cost-benefit analysis.

An important first step towards such an analysis is to examine real estate price fluctuations as a function of noise levels using hedonic regression. \citet{Nelson1982} reviewed early work in this area, but many early studies were limited by small sample sizes and narrow spatial scopes. A review by \citet{Bateman2001} found that the negative eﬀect of traﬃc noise ranged from 0.08 to 2.22 percent per dB with a mean around 0.55 percent per dB, while \citet{Navrud2002} found an average decrease of 0.64 percent per dB. More recent work has tended to focus on the European experience with noise, especially the United Kingdom \citep{Day2007, Blanco2011}, Sweden \citep{Wilhelmsson2000, Andersson2010}, Netherlands \citep{Theebe2004a}, Spain \citep{MarmolejoDuarteCarlos;GonzalezTamez2009}, and Switzerland \citep{Baranzini2010}.

Work examining the negative effects of noise in the United States have tended to concentrate on noise from airports \citep{Espey2000, McMillen2004, Cohen2008a} or indirect effects of traffic noise, such as proximity to highways \citep{Matthews2007, Chernobai2009, Li2012}, or traffic counts \citep{Highes1992, Larsen2012}. the relationship between road traffic noise levels and real estate prices has rarely been explored in the United States, even though the former is perhaps the greatest source of noise in residential neighborhoods (XXX).  One major reason for the thinness of the scholarly literature on this topic in the US lies in the difficulty of modeling noise propagation over the landscape.

To properly analyze the spatial association between real estate prices and exposure to traffic noise at a landscape level, it is necessary to create a noise surface map at sufficiently detailed spatial resolution to account for the complex and heterogeneous interaction between the noise source and the resistance of the landscape to noise propagation.  Implementing such a model can be very difficult.  For one thing, the data needed for the model is very extensive and may not even be readily available (e.g., building footprint and height data).  Furthermore, it is computationally very intensive. Fortunately, recent developments in Geographic Information Systems and distributive computing have reduced these difficulties, making it much easier to create a noise surface map at landscape level.

In this paper, we seek to contribute to the hedonic literature by examining the relationship between real estate prices and variation in road traffic noise exposure.  We approach this issue by means of an empirical case study of the St.\ Paul, Minnesota urban area.  In particular, we construct a locally flexible model that allows for a non-stationary relationship between house prices and our explanatory variables over geography and time. Such Locally Weighted Regression (LWR) models have become increasingly common in published work and have been shown to be more accurate than spatial autoregressive models with misspecified weights matrices.
